---
title: "HIDS 504 Project 1: Infant Birth Weight for First Time Mothers"
author: "Joelle, Junaid, Nelson"
date: "2022-10-18"
output: html_document
---
```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(hexbin)
library(knitr)

opts_chunk$set(echo = FALSE, include=T)
```

# Overview of dataset

### Background
  Low birth-weight (LBW) in infants includes babies who are born preterm (at < 37 weeks) or with fetal growth restriction weighing less than 2500 grams at birth. In the United States (US) and other developed countries, the survival rate of LBW infants has improved over the past 3 decades. However, LBW remains a consistent predictor for infant health, morbidity, and possible demise. Infants born prematurely or with fetal growth restriction are at higher risk for short-term and long-term health morbidities including chronic medical conditions, such as diabetes, hypertension, and heart disease. Associated risk factors for LBW in infants have been noted across maternal populations in the United States including demographic background, socio-economic status, as well as mental and physical health (Goldenberg & Culhane, 2007). Knowing and understanding these risk factors as they are significantly associated with LBW infants may be useful for facilitating interventions and prevention measures to improve pregnancy outcomes. In this exploratory analysis of US birth data gathered from the Centers for Disease Control and Prevention (CDC) in 2018, we explore the relationships between several variables similar to factors shown to affect birth weight in infants against newborn weights reported by the CDC (CDC, 2018).

   An overview of current literature regarding prevalent risk factors for low birth weight in infants revealed maternal demographics, educational background, overall maternal health status and morbidities, prenatal visit frequency, intrauterine infection inducing preterm labor, and smoking history or exposure had a significant impact on pregnancy outcomes and fetal growth (Goldenberg & Culhane, 2007). Especially in the United States, it is observed that African American women, mothers with less education or low socio-economic status have twice the preterm birth rate and higher rates of growth restriction than other maternal populations. According to a cohort study of US births published in 2021, fetal smoking exposure adversely affected fetal growth reflected by lower birth weights in the smoking-exposed population versus non-smoking-exposed newborn population. Additionally, in a prospective hospital based study, researchers observed a high incidence (618 out of 1014 babies) of infants born weighing < 2500 grams in younger mothers age 19-25 and parity zero (having had no children prior).
   
  Knowing this, in our analysis, we chose to consider the data points collected for first-time pregnant mothers in the United States. In this population, we explored if maternal demographics and health status have an impact on infant birth weight. Key variables for our model were chosen based on relevance as seen in current literature and narrowed down excluding those with missing or inappropriate data points, variables that showed high correlation possibly impacting our model results, and categories that did not relate to our chosen population of first-time mothers. The 15 variables chosen include, maternal body mass index (BMI), pregnancy weight gain (WTGAIN), maternal smoking status (CIG_0), interval since last pregnancy as (ILP_R) measured in months, maternal age (MAGER), maternal education (MEDUC), maternal race and hispanic origin (MRAVE6, MHISPX), maternal infection occurrence (NO_INFEC), maternal morbidity and risk assessment (NO_MMORB, NO_RISKS), prenatal visits (PREVIS) and the month prenatal care began (PRECARE), and infant birth weight and sex (DBWT, SEX). It was noted that paternal demographics, education, and health status as well as previous pregnancies may also have an impact on infant birth weight, however, we chose to narrow down our population for the purpose of this analysis to just include data for first-time mothers.
  
  According to the CDC, infants falling within the 5th and 90th average growth rate percentile typically weigh between 2527 grams and 4172 grams. Additionally, the CDC reported mothers should gain 25-35 pounds during pregnancy if they have a normal BMI between 18.5 and 24.9. Mothers who are underweight (BMI less than 18.5) should gain 28-40 pounds and overweight (BMI 25.0-29.9) mothers should gain 15-25 pounds during pregnancy. For anticipated uncomplicated first pregnancies, mothers have prenatal visits every 4 weeks until 28 weeks, every 2 weeks until 36 weeks, and weekly until delivery (CDC, 2022).
  
  Potential confounding variables in our dataset may include maternal mental health status and socioeconomic status. Mental health status of new mothers may result in better or poorer choices throughout pregnancy, thus affecting fetal growth and development. Poor mental health or a negative perspective regarding the current pregnancy may result in less prenatal visits attended due to denial, anxieties, or maternal depression which may delay appropriate and necessary care for both mom and fetus ultimately affecting pregnancy outcomes. Furthermore, lower socioeconomic status may influence a mother’s resources for proper nutrition, basic needs, and healthcare resulting in unhealthy fluctuations in maternal weight gain due to insufficient nutrition or making mothers more prone to morbidities and risk factors associated with low birth weight in infants as discussed above.   

### Load dataset

```{r load_data}
births = read.csv("data/US_births(2018).csv")
```

Base dataset has `r dim(births)[1]` rows (births) and `r dim(births)[2]` columns (variables).

### Key variables

Based on our background research and the area of interest we have chosen to examine,
we have chosen to look more closely at the following variables:

* Mother's pre-pregnancy BMI
* Number of daily cigarettes in 3 months before pregnancy
* Birth weight (in grams): our outcome variable
* Interval since last pregnancy: used to filter data set to only first-time mothers
* Mother's age (in years)
* Mother's education
* Mother's hispanic origin
* Mother's race
* Infections reported during pregnancy (includes gonorrhea, syphillis, chlamydia,
hepatitis B, and hepatitis C)
* Maternal morbidities reported during pregnancy (includes maternal transfusion, 
third or fourth degree perineal laceration, ruptured uterus, unplanned 
hysterectomy, and admission to intensive care unit)
* Risk factors reported during pregnancy (includes diabetes, hypertension, previous
preterm birth, other previous poor pregnancy outcome, pregnancy from infertility
treatment, and previous cesarean delivery)
* Month prenatal care began
* Number of prenatal visits
* Sex of the infant
* Weight gain during pregnancy (in pounds)


### Variables of interest

```{r}
which_vars <- c(
  "BMI",
  "CIG_0",
  "DBWT",
  "ILP_R",
  "MAGER",
  "MEDUC",
  "MHISPX",
  "MRAVE6",
  "NO_INFEC",
  "NO_MMORB",
  "NO_RISKS",
  "PRECARE",
  "PREVIS",
  "SEX",
  "WTGAIN"
)

births_subset <- births[which_vars]
kable(head(births_subset))
```
# Cleaning the Data

### Renaming variables of interest

This is done to make the names more descriptive and easier to read.
```{r}
# renaming vars we're using
new_col_names <- c(
  "mat_bmi_pre",
  "mat_daily_cig_pre",
  "birth_weight_g",
  "months_from_last_preg",
  "mat_age",
  "mat_education",
  "mat_hisp",
  "mat_race",
  "no_infections",
  "no_mat_morbidity",
  "no_risk_factors",
  "month_prenatal_care_began",
  "num_prenatal_visits",
  "sex",
  "mat_weight_gain_lb"
)

colnames(births_subset) <- new_col_names
kable(head(births_subset))
```

### Checking for missing values

Our dataset coded missing values as 9, 99, 999, and 9999 for several columns,
so we need to recode them as NA. Then, we will drop rows with missing values
for any of our variables of interest.
```{r recode_na}
# recode mat_bmi_pre_preg from 99.9 to NA
births_subset$mat_bmi_pre[births_subset$mat_bmi_pre == 99.9] <- NA

# recode mat_cig from 99 to NA
births_subset$mat_daily_cig_pre[births_subset$mat_daily_cig_pre == 99] <- NA

# recode birth_weight_g from 9999 to NA
births_subset$birth_weight_g[births_subset$birth_weight_g == 9999] <- NA

# recode months_from_last_preg from 999 to NA
births_subset$months_from_last_preg[
  births_subset$months_from_last_preg == 999] <- NA

# recode mat_education from 9 to NA
births_subset$mat_education[births_subset$mat_education == 9] <- NA

# recode mat_hisp from 9 to NA
births_subset$mat_hisp[births_subset$mat_hisp == 9] <- NA

# recode mat_race from 9 to NA
births_subset$mat_race[births_subset$mat_race == 9] <- NA

# recode no_infections from 9 to NA
births_subset$no_infections[births_subset$no_infections == 9] <- NA

# recode no_mat_morbidity from 9 to NA
births_subset$no_mat_morbidity[births_subset$no_mat_morbidity == 9] <- NA

# recode no_risk_factors from 9 to NA
births_subset$no_risk_factors[births_subset$no_risk_factors == 9] <- NA

# recode month_prenatal_care_began from 99 to NA
births_subset$month_prenatal_care_began[
  births_subset$month_prenatal_care_began == 99] <- NA

# recode num_prenatal_visits from 99 to NA
births_subset$num_prenatal_visits[
  births_subset$num_prenatal_visits == 99] <- NA

# recode mat_weight_gain_lb from 99 to NA
births_subset$mat_weight_gain_lb[
  births_subset$mat_weight_gain_lb == 99] <- NA
```

```{r missing_val}
kable(colSums(is.na(births_subset)))
```

```{r}
# remove rows with any NAs in subset
births_no_na <- births_subset[complete.cases(births_subset), ]

kable(colSums(is.na(births_no_na)))
```
Our dataset is now free of missing values.

### Recoding categorical vars

Here we re-code the categorical variables in our dataset so that they are factor
variables in R and their values are descriptive.
```{r, echo=T}
# re-coding mat_education
births_no_na$mat_education <- as.factor(births_no_na$mat_education)
births_no_na$mat_education <- recode_factor(
  births_no_na$mat_education,
  "1" = "8th grade or less",
  "2" = "9th through 12th grade",
  "3" = "High school graduate or GED",
  "4" = "Some college, no degree",
  "5" = "Associate degree",
  "6" = "Bachelor's degree",
  "7" = "Master's degree",
  "8" = "Doctorate or Professional degree",
  .ordered=T
)
```

```{r, echo=T}
# re-coding mat_hisp
births_no_na$mat_hisp <- ifelse(births_no_na$mat_hisp == 0, "Non-Hispanic", "Hispanic")

births_no_na$mat_hisp <- as.factor(births_no_na$mat_hisp)
```

```{r, echo=T}
# re-coding no_infections, no_mat_morbidity, no_risk_factors
births_no_na$no_infections <- as.factor(births_no_na$no_infections)
births_no_na$no_infections <- recode_factor(
  births_no_na$no_mat_morbidity, 
  "1" = "No Infections Reported",
  "0" = "Infections Reported"
)

births_no_na$no_mat_morbidity <- as.factor(births_no_na$no_mat_morbidity)
births_no_na$no_mat_morbidity <- recode_factor(
  births_no_na$no_mat_morbidity, 
  "1" = "No Maternal Morbidity Reported",
  "0" = "Maternal Morbidity Reported"
)

births_no_na$no_risk_factors <- as.factor(births_no_na$no_risk_factors)
births_no_na$no_risk_factors <- recode_factor(
  births_no_na$no_risk_factors, 
  "1" = "No Risk Factors Reported",
  "0" = "Risk Factors Reported"
)
```

```{r, echo=T}
# converting sex to factor
births_no_na$sex <- as.factor(births_no_na$sex)
births_no_na$sex <- recode_factor(
  births_no_na$sex, 
  "M" = "Male",
  "F" = "Female"
)
```

### Converting maternal race to dummy variables

We separate maternal race into dummy variables. This is done so that we can use
race as a binary predictor variable since it is an unordered, categorical variable.
```{r, echo=T}
births_no_na$mat_race_white <- as.factor(ifelse(births_no_na$mat_race == 1, 1, 0))
births_no_na$mat_race_black <- as.factor(ifelse(births_no_na$mat_race == 2, 1, 0))
births_no_na$mat_race_aian <- as.factor(ifelse(births_no_na$mat_race == 3, 1, 0))
births_no_na$mat_race_asian <- as.factor(ifelse(births_no_na$mat_race == 4, 1, 0))
births_no_na$mat_race_nhopi <- as.factor(ifelse(births_no_na$mat_race == 5, 1, 0))
births_no_na$mat_race_multiple <- as.factor(ifelse(births_no_na$mat_race == 6, 1, 0))
```

### Filter for only first pregnancies

Now, we finally subset our data so that it includes our group of interest,
births for first-time mothers.
```{r, echo=T}
# select first pregnancies
births_clean <- births_no_na[births_no_na$months_from_last_preg == 888, ]

# drop months_from_last_preg
births_clean <- subset(births_clean, select = -c(months_from_last_preg))
```

Our clean dataset has `r dim(births_clean)[1]` rows and `r dim(births_clean)[2]` columns.

```{r}
kable(head(births_clean))
```


### Saving Cleaned Dataset

Output our clean dataset as a CSV.
```{r}
write.csv(births_clean, "data/births_clean.csv", row.names=F)
```

# Exploratory Data Analysis

Here we explore some summary statistics and plot our variables of interest so
that we get a better idea of what the data looks like.

### Checking normality of birth weight

```{r table_birth_weight_g}
summary(births_clean$birth_weight_g)
```

```{r hist_birth_weight_g}
BINWIDTH <- 100

ggplot(births_clean, aes(birth_weight_g)) +
  geom_histogram(binwidth = BINWIDTH, fill="dark blue") +
  geom_density(aes(y = BINWIDTH * ..count..), color="red") +
  xlab("Birth Weight (grams)") +
  ylab("Count") +
  ggtitle("Birth Weight (grams)") +
  theme(plot.title = element_text(hjust = 0.5))
```

Since birth weight already appears to follow a normal distribution, we don't 
need to perform a log2 transformation. The median birth weight is 3265g.

##### Mother's BMI Pre-Pregnancy

```{r table_mat_bmi_pre}
summary(births_clean$mat_bmi_pre)
```

```{r birth_weight_mat_bmi_pre}
ggplot(births_clean, aes(x = mat_bmi_pre, y = birth_weight_g)) + 
  geom_hex(bins=60) +
  geom_smooth(method="lm", color="green") +
  xlab("Mother's BMI Pre-Pregnancy") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Mother's Pre-Pregnancy BMI") +
  theme(plot.title = element_text(hjust = 0.5))
```

The median BMI of first-time mother's pre-pregnancy is 24.8. There appears to be
a slight positive correlation between a mother's pre-pregnancy BMI and infant
birth weight.

##### Mother's Daily Cigarette Use Pre-Pregnancy

```{r table_mat_daily_cig_pre}
summary(births_clean$mat_daily_cig_pre)
```


```{r hex_mat_daily_cig_pre}
ggplot(births_clean, aes(x = mat_daily_cig_pre, y = birth_weight_g)) + 
  geom_hex(bins=60) +
  geom_smooth(method="lm", color="green") +
  xlab("Mother's Daily Cigarette's Pre-Pregnancy") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Mother's Pre-Pregnancy Cigarettes") +
  theme(plot.title = element_text(hjust = 0.5))
```

The median daily cigarettes consumed for mothers pre-pregnancy was 0.
There appears to be a slight negative correlation between a mother's 
pre-pregnancy cigarette use and infant birth weight.

##### Mother's Age

```{r table_mat_age}
summary(births_clean$mat_age)
```

```{r hex_mat_age}
ggplot(births_clean, aes(x = mat_age, y = birth_weight_g)) + 
  geom_hex(bins=30) +
  geom_smooth(method="lm", color="green") +
  xlab("Mother's Age (years)") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Mother's Age") +
  theme(plot.title = element_text(hjust = 0.5))
```

Median age for the first-time mothers in our dataset was 26 years old. Ages 
ranged from 12 to 50 years old. There doesn't appear to be much, if any, 
correlation between maternal age and infant birth weight.

##### Mother's Education

```{r table_mat_education}
tapply(births_clean$birth_weight_g, births_clean$mat_education, summary)
```

```{r violin_mat_education}
ggplot(births_clean, aes(x=mat_education, y=birth_weight_g, fill=mat_education)) + 
  geom_violin() +
  geom_boxplot(color = "grey", alpha = 0.2, width = 0.2, outlier.shape = NA, show.legend = F) +
  xlab("Mother's Education") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Mother's Education") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
        legend.title = element_blank(), plot.title = element_text(hjust = 0.5))
```

Mother's who had obtained a bachelor's degree had the highest median infant 
birth weight at 3317g. The lowest was for those who had completed 9th-12th 
grade at 3180g. Overall the values between education groups appear to be similar
and they appear to follow similar distributions.

##### Mother's Hispanic Origin

```{r table_mat_hisp}
tapply(births_clean$birth_weight_g, births_clean$mat_hisp, summary)
```

```{r violin_mat_hisp}
ggplot(births_clean, aes(x=mat_hisp, y=birth_weight_g, fill=mat_hisp)) + 
  geom_violin(show.legend = F) +
  geom_boxplot(color = "black", alpha = 0.2, width = 0.2, outlier.shape=NA, show.legend = F) +
  xlab("Mother's Hispanic Origin") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Mother's Hispanic Origin") +
  theme(legend.title = element_blank(), plot.title = element_text(hjust = 0.5))
```

The median infant birth weight for non-hispanic mothers (3274g) is slightly 
higher than that for hispanic mothers (3250g). They appear to follow similar
distributions.

##### Mother's Race

```{r table_mat_race}
tapply(births_clean$birth_weight_g, births_clean$mat_race, summary)
```

```{r violin_mat_race}
ggplot(births_clean, aes(x=as.factor(mat_race), y=birth_weight_g, fill=as.factor(mat_race))) + 
  geom_violin() +
  geom_boxplot(color = "black", alpha = 0.2, width = 0.2, outlier.shape=NA, show.legend = F) +
  xlab("Mother's Race") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Mother's Race") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
        legend.title = element_blank(), plot.title = element_text(hjust = 0.5)) +
  scale_fill_discrete(labels = c("White", "Black", "AIAN", "Asian", "NHOPI", "Multiple"))
```

White and American Indian/Alaska Native mothers had the highest median infant
birth weights at 3310g. Black mothers had the lowest at 3118g. 

##### Mother's Reported Infections

```{r table_no_infections}
tapply(births_clean$birth_weight_g, births_clean$no_infections, summary)
```

```{r violin_no_infections}
ggplot(births_clean, aes(x=no_infections, y=birth_weight_g, fill=no_infections)) + 
  geom_violin(show.legend = F) +
  geom_boxplot(color = "black", alpha = 0.2, width = 0.2, outlier.shape=NA, show.legend = F) +
  xlab("Infections") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Infections Reported") +
  theme(legend.title = element_blank(), plot.title = element_text(hjust = 0.5))
```

Median birth weights for mothers who reported infections during their pregnancy
were higher at 3420g than those who did not at 3260g.

##### Mother's Reported Morbidities

```{r table_no_mat_morbidity}
tapply(births_clean$birth_weight_g, births_clean$no_mat_morbidity, summary)
```

```{r violin_no_mat_morbidity}
ggplot(births_clean, aes(x=no_mat_morbidity, y=birth_weight_g, fill=no_mat_morbidity)) + 
  geom_violin(show.legend = F) +
  geom_boxplot(color = "black", alpha = 0.2, width = 0.2, outlier.shape=NA, show.legend = F) +
  xlab("Maternal Morbidity") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Maternal Morbidity Reported") +
  theme(legend.title = element_blank(), plot.title = element_text(hjust = 0.5))
```

Median birth weights for mothers who reported morbidity during their pregnancy
were higher at 3420g than those who did not at 3260g.

##### Mother's Reported Risk Factors

```{r table_no_risk_factors}
tapply(births_clean$birth_weight_g, births_clean$no_risk_factors, summary)
```

```{r violin_no_risk_factors}
ggplot(births_clean, aes(x=no_risk_factors, y=birth_weight_g, fill=no_risk_factors)) + 
  geom_violin(show.legend = F) +
  geom_boxplot(color = "black", alpha = 0.2, width = 0.2, outlier.shape=NA, show.legend = F) +
  xlab("Risk Factors") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Risk Factors Reported") +
  theme(legend.title = element_blank(), plot.title = element_text(hjust = 0.5))
```

Median birth weights for mothers who reported risk factors during their pregnancy
were lower at 3160g than those who did not at 3289g.

##### Month Prenatal Care Began

```{r table_month_prenatal_care_began}
summary(births_clean$month_prenatal_care_began)
```

```{r hex_month_prenatal_care_began}
ggplot(births_clean, aes(x = month_prenatal_care_began, y = birth_weight_g)) + 
  geom_hex(bins=30) +
  geom_smooth(method="lm", color="green") +
  scale_x_continuous(n.breaks=11) +
  xlab("Month of Pregnancy Prenatal Care Began") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Month Prenatal Care Began") +
  theme(plot.title = element_text(hjust = 0.5))
```

Most mothers began prenatal care in month 2 or 3 of their pregnancy. There
doesn't appear to be much correlation between the month care began and 
infant birth weight.

##### Number of Prenatal Visits

```{r table_num_prenatal_visits}
summary(births_clean$num_prenatal_visits)
```

```{r hex_num_prenatal_visits}
ggplot(births_clean, aes(x = num_prenatal_visits, y = birth_weight_g)) + 
  geom_hex(bins=60) +
  geom_smooth(method="lm", color="green") +
  xlab("Number of Prenatal Visits") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Number of Prenatal Visits") +
  theme(plot.title = element_text(hjust = 0.5))
```

The median number of prenatal visits was  12. The number of prenatal visits 
appears to be positively correlated with birth weight.

##### Sex

```{r table_sex}
tapply(births_clean$birth_weight_g, births_clean$sex, summary)
```

```{r violin_sex}
ggplot(births_clean, aes(x=sex, y=birth_weight_g, fill=sex)) + 
  geom_violin(show.legend = F) +
  geom_boxplot(color = "black", alpha = 0.2, width = 0.2, outlier.shape=NA, show.legend = F) +
  xlab("Sex") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Sex") +
  theme(legend.title = element_blank(), plot.title = element_text(hjust = 0.5))
```

Male infant birth weight has a median of 3320g, slightly higher than that of 
female infants (3212g). Overall they follow similar distributions.

##### Mother's Weight Gain During Pregnancy

```{r birth_weight_mat_weight_gain_lb}
summary(births_clean$mat_weight_gain_lb)
```

```{r hex_mat_weight_gain_lb, echo=F}
ggplot(births_clean, aes(x = mat_weight_gain_lb, y = birth_weight_g)) + 
  geom_hex() +
  geom_smooth(method="lm", color="green") +
  xlab("Mother's Weight Gain During Pregnancy") +
  ylab("Birth Weight (grams)") +
  ggtitle("Birth Weight by Mother's Weight Gain During Pregnancy") +
  theme(plot.title = element_text(hjust = 0.5))
```

The average first-time mother in our dataset gained around 31 lbs over their 
pregnancy. Mother's weight gain appears to be positively correlated with 
the infant's birth weight.
# Variable Correlations

```{r var_correlation}

library(corrplot)

births_clean_no_num <- subset(births_clean , select = -c(sex , mat_education , mat_race_white,mat_race_aian,mat_race_nhopi, mat_race_black, mat_race_asian, mat_race_multiple, no_infections, mat_hisp,no_mat_morbidity ,no_risk_factors ))

corrplot(cor(births_clean_no_num),method='color'  , order = 'alphabet' , type='lower', cl.ratio=0.2 , tl.cex =0.8 , tl.srt=45 , tl.col = 'black' ,col = COL2('RdBu', 10) , title = 'Correlation Matrix' , mar=c(0,0,1,0))

```

# Hypothesis testing

## Explanation of hypothesis & parameters

Null hypothesis: There is no relationship between our independent variables
or interest and the birth weight of the infant.

Parameters: 

* Birth weight (in grams): our dependent/outcome variable
* Mother's pre-pregnancy BMI
* Mother's cigarette use in 3 months before pregnancy 
* Mother's age (in years)
* Mother's education
* Mother's Hispanic origin
* Mother's race
* Infections reported during pregnancy
* Morbidity reported during pregnancy
* Risk factors reported during pregnancy
* Month prenatal care began
* Number of prenatal visits
* Sex of the infant
* Weight gain during pregnancy (in pounds)

## Explanation of model

We will use a multiple linear regression model to estimate the relationship 
between our outcome variable, birth weight of the infant, and our independent
variables. This model will attempt to model a linear relationship between these.



### Select Columns for regression
```{r prep_run}
which_vars_x<- colnames(births_clean)[!colnames(births_clean) %in% c("birth_weight_g"  , "mat_race" ,"no_mat_morbidity" , "mat_race_multiple")]
x_vars_str = paste(which_vars_x, collapse = " + ")
form_str = paste("birth_weight_g ~ ", x_vars_str, sep="")
form = as.formula(form_str)
```

### Run Linear Regression

```{r run_test}
model <- lm(formula=form, data = births_clean)
```

# Results/Conclusions

### Summary Statistics of the model

```{r summ_multi}
library(jtools)
summ(model)
```


The model's output is summarized above. The thing to note is that P-values for every variable except mat_race_nhopi show high levels of significance. However the adjusted R2 value is 0.13. Hence the model is not very effective at predictions.

Comparing the coefficients, no_risk_factor has the biggest change on birth_weight_g. Mothers with associated risk factors had babies 215grams lighter than mother without.

Female babies were also lighter than male babies by 99grams


### Confidence Intervals
```{r confint}
confint(model)
```
The confidence Interval for the variables are listed above

## Run Single Linear Regression for some variables

Based on the exploratory data analysis we run single variable linear regression on a few of the independent variables. The results can be seen in the following section

```{r run_single_test}

model_bmi <- lm(birth_weight_g ~ mat_bmi_pre, data = births_clean)
model_cig <- lm(birth_weight_g ~ mat_daily_cig_pre, data = births_clean)
model_ed <- lm(birth_weight_g ~ mat_education, data = births_clean)
model_visits <- lm(birth_weight_g ~ num_prenatal_visits, data = births_clean)
model_weight <- lm(birth_weight_g ~ mat_weight_gain_lb, data = births_clean)


```
### Summary Statistics for single linear regression

```{r sum_single}
library(jtools)
summ(model_bmi)
summ(model_cig)
summ(model_ed)
summ(model_visits)
summ(model_weight)


```

The results for the linear regression run on variables mat_pre_bmi, mat_daily_cig, mat_education, num_prenatal_visits, mat_weight_gain_lb can be seen above. Individually the variables have an even lower R2 value. 

# Conclusion

The multiple linear regression model that we developed has a very low R2 score of 0.13. The variablesa are significant but the model is not effective at predicting the outcome. We can try and gather more variables which hopefully can explain the variations in our data. Lastly we cannot rule out the fact that there is an inherently high amount of unxplainable variabilty in the dataset.






# Sources

United States, Department of Health and Human Services, Centers for Disease Control and Prevention. “User Guide to the 2018 Natality Public Use File”. Centers for Disease Control and Prevention, 2018.

United States, Department of Health and Human Services, Centers for Disease Control and Prevention. “Weight Gain during Pregnancy.” Centers for Disease Control and Prevention, Centers for Disease Control and Prevention, 13 June 2022, https://www.cdc.gov/reproductivehealth/maternalinfanthealth/pregnancy-weight-gain.htm.

Xu, Richard et al. “DNA methylation mediates the effect of maternal smoking on offspring birthweight: a birth cohort study of multi-ethnic US mother-newborn pairs.” Clinical epigenetics vol. 13,1 47. 4 Mar. 2021, doi:10.1186/s13148-021-01032-6
 
Robert L Goldenberg, Jennifer F Culhane, Low birth weight in the United States, The American Journal of Clinical Nutrition, Volume 85, Issue 2, February 2007, Pages 584S–590S, https://doi.org/10.1093/ajcn/85.2.584S
 
Metgud, Chandra S et al. “Factors affecting birth weight of a newborn--a community based study in rural Karnataka, India.” PloS one vol. 7,7 (2012): e40040. doi:10.1371/journal.pone.0040040
 
Raman, Ts Raghu et al. “Low Birth Weight Babies: Incidence and Risk Factors” Medical journal, Armed Forces India vol. 54,3 (1998): 191-195. doi:10.1016/S0377-1237(17)30539-7

Carter, E B et al. “Number of prenatal visits and pregnancy outcomes in low-risk women.” Journal of perinatology : official journal of the California Perinatal Association vol. 36,3 (2016): 178-81. doi:10.1038/jp.2015.183

```{r}
sessionInfo()
```



