---
title: "Project 1"
author: "Joelle, Junaid, Nelson"
date: "2022-10-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=TRUE, fig.path="figures/")

library(ggplot2)
```

# Overview of dataset

### Background



https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3390317/
- mean birth weight: 2.6kg (range: [1.2kg, 3.8kg])
- significant risk factors: maternal education, smoking exposure, age at first
  pregnancy, birth interval, previous history of low birth weight baby, weight
  gain, maternal weight at last week of gestation, pregnancy induced hypertension,
  high risk pregnancy, late antenatal registration
  
https://academic.oup.com/ajcn/article/85/2/584S/4649824
- income, race, ethnicity, infection, maternal BMI

https://www.jkscience.org/archive/Volume81/epidem.pdf
- avg bw: 2.67 +- 0.42kg
- antenatal care, parity, inter-pregnancy interval, gestational weight, OB history

https://www.hindawi.com/journals/jp/2013/172395/
- avg bw 2006: 2950.8g
- avg bw 2010: 2937.5g
- # of prev. preg., smoking, gender, mat. age, mat. height, weight gain during
  preg., BMI, use of in vitro, induction of labor, gestational duration

http://library.isical.ac.in:8080/jspui/bitstream/10263/2925/1/Binder1.pdf
- infant sex, mat. age, mat. edu., mat. ethnicity, pat. income, mat. nutrition,
  antenatal visits, 
- Low BW defined by WHO as < 2500g (maybe something to mention in background, 
  although probably not part of our actual analysis, can talk about consequences
  of low bw for infants?)

### Load dataset

```{r}
births = read.csv("code/US_births(2018).csv")
```

# Key variables

### Variables key 
#### (probably just need this for the vars we're keeping, just including the whole
#### thing for now as we choose - my initial thoughts of what to keep have a +
#### and what to drop have a - , outcome has +++:) 
-ATTEND:Attendant at birth
  1 Doctor of Medicine (MD)
  2 Doctor of Osteopathy (DO)
  3 Certified Nurse Midwife (CNM)
  4 Other Midwife
  5 Other
  9 Unknown or not stated)
-BFACIL: Birth Place
  1 Hospital
  2 Freestanding Birth Center
  3 Home (intended)
  4 Home (not intended)
  5 Home (unknown if intended)
  6 Clinic / Doctor’s Office
  7 Other
  9 Unknown
+BMI: Body Mass Index
  13.0-69.9 Body Mass Index
  99.9 Unknown or not stated
+CIG_0: Cigarettes Before Pregnancy
  00-97 Number of cigarettes daily
  98 98 or more cigarettes daily
  99 Unknown or not stated
+++DBWT: Birth Weight – Detail in Grams (Edited)
  0227-8165 Number of grams
  9999 Not stated birth weight
-DLMP_MM: Last Normal Menses Month
  01 January
  02 February
  03 March
  04 April
  05 May
  06 June
  07 July
  08 August
  09 September
  10 October
  11 November
  12 December
  99 Unknown or not stated
-DLMP_YY: Last Normal Menses Year
  nnnn Year of last normal menses
  9999 Unknown or not stated
  (maybe use M/Y to calc. time since last normal menses? or maybe not relevant, idk)
-DMAR: Marital Status
  1 Married
  2 Unmarried
-DOB_MM: Birth Month
-DOB_TT: Time of Birth
  0000-2359 Time of Birth
  9999 Not Stated
-DOB_WK: Birth Day of Week
  1 Sunday to 7 Saturday
-DOB_YY: Year of birth (2018?)
-DWgt_R: Delivery Weight Recode
  100-400 Weight in pounds
  999 Unknown or not stated
-FAGECOMB: Father's Combined Age
  09-98 Father's combined age in years
  99 Unknown or not stated
+FEDUC: Father's Education
+FHISPX: Father's Hispanic Origin
-FRACE15: Father's Race 
  (15 options)
-FRACE31: Father's Race
  31 options + 99 for unknown/not stated
+FRACE6: Father's Race
  1 White (only)
  2 Black (only)
  3 AIAN (only)
  4 Asian (only)
  5 NHOPI (only)
  6 More than one race
  9 Unknown or Not Stated
-ILLB_R: Interval Since Last Live Birth
  000-003 Plural delivery
  004-300 Months since last live birth
  888 Not applicable / 1st live birth
  999 Unknown or not stated
-ILOP_R: Interval Since Last Other Pregnancy
  000-003 Plural delivery
  004-300 Months since last live birth
  888 Not applicable / 1st natality event
  999 Unknown or not stated
+ILP_R: Interval Since Last Pregnancy
  000-003 Plural delivery
  004-300 Months since last live birth
  888 Not applicable / 1st natality event
  999 Unknown or not stated
-IMP_SEX: Imputed Sex
  Blank Infant Sex not Imputed
  1 Infant Sex is Imputed
-IP_GON: Infections Present - Gonorrhea
  Y Yes
  N No
  U Unknown or not stated
-LD_INDL: Induction of Labor
  Y Yes
  N No
  U Unknown or not stated
+MAGER: Mother's Single Years of Age
  12 10-12 years
  n = 12-49 n years
  50 50 years and over
-MAGE_IMPFLG: Mother's Age Imputed
-MAR_IMP: Mother's Marital Status Imputed
-MBSTATE_REC: Mother's Nativity
  1 Born in the U.S. (50 US States)
  2 Born outside the U.S. (includes possessions)
  3 Unknown or Not Stated
+MEDUC: Mother's Education
+MHISPX: Mother's Hispanic Origin
-MM_AICU: Admit to Intensive Care
  Y Yes
  N No
  U Unknown or not stated
-MRACE15: Mother's Race
-MRACE31: Mother's Race
-MRACEIMP: Mother's Race Imputed
+MRAVE6: Mother's Race
-MTRAN: Mother Transferred
  Y Yes
  N No
  U Unknown or not stated  
-M_Ht_In: Mother's Height in Total Inches
  30-78 Height in inches
  99 Unknown or not stated
+NO_INFEC: No Infections Reported
  1 True
  0 False
  9 Not Reported
+NO_MMORB: No Maternal Morbidity Reported
  1 True
  0 False
  9 Not Reported
+NO_RISKS: No Risk Factors Reported
  1 True
  0 False
  9 Not Reported
-PAY: Payment Source for Delivery
  1 Medicaid
  2 Private Insurance
  3 Self-Pay
  4 Indian Health Service
  5 CHAMPUS/TRICARE
  6 Other Government (Federal, State, Local)
  8 Other
  9 Unknown
-PAY_REC: Payment Recode
  1 Medicaid
  2 Private Insurance
  3 Self Pay
  4 Other
  9 Unknown
+PRECARE: Month Prenatal Care Began
  00 No prenatal care
  01-10 Month prenatal care began
  99 Unknown or not stated
+PREVIS: Number of Prenatal Visits
  00-98 Number of prenatal visits
  99 Unknown or not stated
+PRIORDEAD: Prior Births Now Dead
  00-30 Number of children dead from previous live births.
  99 Unknown or not stated
+PRIORLIVE: Prior Births Now Living
  00-30 Number of children still living from previous live births.
  99 Unknown or not stated
+PRIORTERM: Prior Other Terminations
  00-30 Number other terminations
  99 Unknown or not stated
+PWgt_R: Pre-pregnancy Weight Recode
  075-375 Weight in pounds
  999 Unknown or not stated
-RDMETH_REC: Delivery Method Recode
  1 Vaginal (excludes vaginal after previous C-section)
  2 Vaginal after previous c-section
  3 Primary C-section
  4 Repeat C-section
  5 Vaginal (unknown if previous c-section)
  6 C-section (unknown if previous c-section)
  9 Not stated
-RESTATUS: Residence Status
-RF_CESAR: Previous Cesarean
  Y Yes
  N No
  U Unknown or not stated
-RF_CESARN: Number of Previous Cesareans
  00 None
  01-30 Number of previous cesareans
  99 Unknown or not stated
+SEX: Sex of Infant
  M Male
  F Female
+WTGAIN: Weight Gain
  00-97 Weight gain in pounds
  98 98 pounds and over
  99 Unknown or not stated

### Variables of interest

```{r}
vars_keep <- c(
  "BMI",
  "CIG_0",
  "DBWT",
  "FEDUC",
  "FHISPX",
  "FRACE6",
  "ILP_R",
  "MAGER",
  "MEDUC",
  "MHISPX",
  "MRAVE6",
  "NO_INFEC",
  "NO_MMORB",
  "NO_RISKS",
  "PRECARE",
  "PREVIS",
  "PRIORDEAD",
  "PRIORLIVE",
  "PRIORTERM",
  "PWgt_R",
  "SEX",
  "WTGAIN"
)

births_subset <- births[vars_keep]
head(births_subset)
```



```{r}
# renaming vars we're using?

```

# Exploratory data anlysis

### Checking for missing values

```{r missing_val}
# this doesn't capture everything, they used 99, 999, & 9999 in some cols for NAs
knitr::kable(colSums(is.na(births)))
```

### Checking normality

```{r birth_weight_hist}
BINWIDTH <- 100

ggplot(births, aes(DBWT)) +
  geom_histogram(binwidth = BINWIDTH) +
  geom_density(aes(y = BINWIDTH * ..count..), color="red")
```
```{r birth_weight_stats}
summary(births$DBWT)
```

```{r log_birth_weight_hist}
ggplot(births, aes(log2(DBWT))) +
  geom_histogram(aes(y = after_stat(density))) +
  geom_density(color = "red")
```

### Tables & plots, etc.

```{r table_1}

```

```{r plot_1}

```

# Hypothesis testing

### Explanation of hypothesis & parameters

### Explanation of model (multiple linear regression)

### Running the test

```{r}
x_vars = paste(vars_keep, collapse = " + ")
form_str = paste("DBWT ~ ", x_vars, sep="")
form = as.formula(form_str)
```

```{r}
lm(formula=form, data = births_subset)
```
```{r}
summary(lm(formula=form, data = births_subset))
```

### Results/Conclusions Interpretation

```{r}
sessionInfo()
```





